\documentclass{article}
\usepackage{shortex}
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\usepackage{xcolor}
% \usepackage{jupynotex}
\usepackage{csvsimple}
\usepackage{placeins}
% \usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}


\graphicspath{{./graphics}}

\lstdefinelanguage{Julia}{
    keywords={function, end, if, else, elseif, for, while, return, 
               local, global, const, struct, mutable, abstract, type,
               using, import, export, module, begin, let, do, try, 
               catch, finally, true, false, nothing},
    sensitive=true,
    comment=[l]{\#},
    morecomment=[s]{\#=}{=\#},
    morestring=[b]",
    morestring=[b]',
}

\lstset{
    language=Julia,
    basicstyle=\ttfamily\small,
    keywordstyle=\color{blue}\bfseries,
    commentstyle=\color{gray}\itshape,
    stringstyle=\color{purple},
    numberstyle=\tiny\color{gray},
    numbers=left,
    stepnumber=1,
    frame=single,
    breaklines=true,
    captionpos=b,
    tabsize=4,
    showstringspaces=false,
}

\lstset{
    literate={ρ}{$\rho$}{1}
             {α}{$\alpha$}{1}
             {β}{$\beta$}{1}
             {σ}{$\sigma$}{1}
             {σ²}{$\sigma^2$}{2}
             {μ}{$\mu$}{1}
             {ϵ}{$\epsilon$}{1}
             {η}{$\eta$}{1}
             {η²}{{$\eta^2$}}{2}
             {r̂}{$\hat r$}{1}
             {Δ}{$\Delta$}{1}
             {Λ}{$\Lambda$}{1}
             {∞}{$\infty$}{1}
}

\begin{document}
\title{STAT 547E - Report - Diffusion Schrödinger Bridge}
\author{Zachary Lau}
\date{\today}
\maketitle

\begin{abstract}
The Schrödinger Bridge problem is a classic problem in physics, which has
received renewed attention in statistics thanks to its application to diffusion
models. This report provides an accesible introduction to the Schrödinger bridge
problem for statisticians, discusses its connection to diffusion models, and
provides a simple MCMC algorithm to sample from a finite particle approximation.
\end{abstract}

\section{Introduction}
Recent work in generative modelling has brought particular focus to the area
of diffusion processes
\cite{ho2020denoisingdiffusionprobabilisticmodels}
\cite{song2020generativemodelingestimatinggradients}. Amongst these works,
\cite{debortoli2023diffusionschrodingerbridgeapplications}
makes use of the concept of a Schrödinger bridge
\cite{Schrodinger1932}; a diffusion process with
constrained marginals that ensures finite time transitions from the data
distribution to the prior.

This report builds up intuition about the Schrödinger bridge problem from the
ground up, based on a finite particle perspective.
We begin with discussing Brownian bridges, the fundamental building blocks of
the Schrödinger bridge.  We then move onto mixtures of Brownian bridges which
arise when we have multiple pairs of particles. By allowing the particles'
pairings to be chosen probabilistically according to the diffusion dynamics, we
recover a finite version of the Schrödinger bridge problem. In the limit of
infinite particles, this approaches exactly the standard Schrödinger bridge
problem through one-dimensional visualizations.

In addition to introducing the problem formulation, we discuss the classical
Sinkhorn solution, as well as the novel generative modelling centered
formulation provided by DeBortoli et al
\cite{debortoli2023diffusionschrodingerbridgeapplications}.  Finally we
introduce a simple MCMC algorithm to simulate from the finite particle system,
and graphically demonstrate the simulations consistency with the 
Schrödinger bridge problem.

\section{Brownian Bridge}
The Brownian Bridge can be seen as the precursor problem of the Schrödinger
Bridge Problem. The set up is as follows.  Consider a particle that begins at point $X_0
= 0$ at time $t=0$. This particle evolves according to a standard Brownian
motion (e.g. Wiener process with unit variance). Then conditioning on the
starting position, the location of the particle at time $t$ is given by 
\begin{align}
    X_t|X_0 = 0 \sim \Norm(0, t)
\end{align}
Now additionally consider observing $X_1 = 0$. Conditioning on both $X_0 = 0$
and $X_1 = 0$, the distribution of $X_t$ can be found analytically. Based on the
Markovian property of the Brownian motion, $X_1 | X_t \sim \Norm(X_t, 1-t)$.
In other words $X_1 = X_t - (1-t) Z$ where $Z \sim \Norm(0,1)$ is draw
independently of $X_t$. This gives $\Cov(X_t, X_1) = \Var(X_t) = t$, and hence
we recover the joint distribution 
\begin{align}
    X_t, X_1 | X_0 = 0 \sim \Norm\left( 
        \begin{bmatrix} 0 \\ 0 \end{bmatrix},
        \begin{bmatrix}
            t & t \\
            t & 1
        \end{bmatrix}
    \right)
\end{align}
Now conditioning on $X_1 = 0$ as well, we find
\begin{align}
    X_t | X_1 = 0, X_t = 0 \sim \Norm(0, t(1-t))
\end{align}
Here we derive the variance from the Schur complement, or equivalently simply
from the properties of multivariate Gaussians. Considering all $t$
simultaneously, this is known as a Brownian bridge. This result is quite
aesthetic, we note that it is symmetric, which reflects the symmetry in Brownian
motion.  Furthermore as Schrödinger points out in his original paper, the
density can be seen as a \emph{product} of a density term from a Brownian motion
starting at $X_0=0$ and another time-reversed Browniain motion starting at
$X_1=0$. As a minor extension, if we instead consider a particle starting at
$X_0=a$ and ending at $X_1=b$, we get the intermediary distributions
\begin{align}
    X_t | X_0 = a, X_1 = b \sim \Norm(ta + (1-t)b, t(1-t))
\end{align}
In other words, we have simply added a linear drift term. 
% TODO - do we talk about the corresponding SDE?

\section{Brownian Bridges - Multiple Independent Particles}
We next consider a system of $N$ particles, each independently undergoing
Brownian motion. Suppose that we know the starting and ending point of each
particle. We use superscripts to denote particle indices. That is we know
$X^1_0 = a_1, X^2_0 = a_2, \ldots, X^1_1 = b_1, X^2_1 = b_2, \ldots$. Each
particle independently traverses a Brownian bridge from its starting point to
its end point. That is
\begin{align}
    X^i_t | X^i_0 = a_i, X^i_1 = b_i \sim \Norm(ta_i + (1-t)b_i, t(1-t))
\end{align}
Looking at the particles as a collective, we might be
concerned with the overall \emph{density} of particles at time $t$. This 
density is given by a \emph{mixture of normals} defined by the respective
Brownian Bridges.

\section{Unknown Pairings - A Bayesian Perspective}
Next we will generalize further. Suppose that we observe the locations of the
$N$ particles at the beginning and the end, but we don't know which particle
correponds to which! How can we pair the particles? We will take an approach
that is a hybrid of statistical mechanics and Bayesian inference. First we place
a \emph{uniform prior} on all possible pairings. Then we consider conditioning
on the \emph{difference} between the start and the end points, that is we 
condition on $X^i_1 - X^i_0$. We note that we have to work with differences
because we haven't fully specified a probability model; the Brownian motion only
specifies the relationship between $X_1$ and $X_0$. This gives a Gaussian
likelihood $\mathcal L \propto \exp\left(-\frac 1 2(x_1-x_0)^2\right)$. Denoting
the pairing by the tuple of vectors $(\mathbf i, \mathbf j)$ and, 
taking this product over all $i$ points, we get the posterior over pairings by
\begin{align}
    p(\mathbf i,\mathbf j) \propto \prod_{(i,j)} \exp\left(-\frac 1 2(a_i-b_j)^2\right)
\end{align}
(To ensure the pairing is uniquely identifiable we may fix $\mathbf i = [1, 2,
3, \ldots, N]$). We make an interesting physical remark that will become
important later. Supposing that we associated $q$ with the counting measure on
$[N] \times [N]$ where $[n] = \{1, \ldots N\}$ such that $q(i,j)=\frac 1 N$ if
the particle that starts at $i$ ends up at location $j$ and $0$ otherwise (i.e.
it absorbs all the information about which particle goes where). Let
$r$ be the reference measure on $[N] \times [N]$ given by
$r(i,j) = \exp\left(-\frac 1 2(a_i-b_j)^2\right)$. Then $r$ contains all the
information about how Brownian motion actually works (i.e. about the underlying
system). It is typically not noramlized. The log posterior turns out to be the
negative cross-entropy between $q$ and $r$. That is
\begin{align}
    -H(q,r) &= \int \log r(i,j) \d q \\
    &= \frac 1 N \sum_{(i,j)} -\frac 1 2 (a_i - b_j)^2 \\
    &= \frac 1 N \log p(\mathbf i, \mathbf j) \\
    \log p(\mathbf i, \mathbf j) &= -NH(q,r)
\end{align}
Already, we can see some interesting phenomena. Firstly, we notice the
similarity beween the sum of squares losses and the cost function in the
Monge-Kantorovich problem. We will return to this later. Furthermore, notice
that multiplying the log-likelihood by $N$ corresponds to a \emph{power
posterior}. Therefore for large $N$ we might expect the posterior to concentrate
near the minimizer of the cross-entropy.

\section{Many particles, finite points}
\label{sec:many_particles}
We generalize once more. Let us consider $M$ particles starting at each of
the $N$ locations, for a total of $M \times N$ particles. We need to keep
track of how many particles from each of the $N$ starting locations end up in 
each of the $N$ ending locations. We can easily generalize the discrete measure
$q$ from before to such that $q(i,j)$ is the \emph{fraction of particles} starting
at location $i$ and ending up at location $j$. However, now we have to be a
little bit more careful with how we pick our prior. This gets a little confusing
to think of priors over measures, so we introduce another way to think of 
$q$-as a matrix. In this case let the matrix be given by $Q_{i,j} = q(i,j)$.

Firstly we must constrain the support such that the each row and column of 
$Q$ sums to $\frac 1 N$. These are our \emph{marginal} constraints. We next must
consider actually assigning matches between particles on the left and the right.
We again assume a uniform distribution over all possible pairings that respect
our marginal constraints. Note there is a distinction between a uniform 
distribution on pairings and uniform distribution on $Q$
matrices. For example if we swap the labels on two particles we have a new
pairing, but the $Q$ matrix is unchanged! Indeed we have discovered the
difference between a microstate (pairing) and macro-state ($Q$-matrix).

For our $Q$ matrix this has \emph{combinatorial}
implications. The total number of pairings correponding to a $Q$ matrix is
given by a multinomial, namely
\begin{align}
    \frac{(NM)!}{\prod_{i,j} (NMQ_{i,j})!}
\end{align}
Taking into account
the Brownian Noise likelihood from early, we get the posterior
\begin{align}
    p(Q) \propto \frac{1}{\prod_{i,j} (NMQ_{i,j})!}\prod_{(i,j)} \exp\left(-\frac 1 2(a_i-b_j)^2\right) \\
    \log p(Q) = - \sum_{i,j} \log (NMQ_{i,j})! + \sum_{(i,j)} - \frac 1 2 NMQ_{i,j}(a_i - b_j)^2 + \log Z
\end{align}
Where $Z$ is a normalizing constant. Viewing $Q$ as a probability measure, and using $r$
to denote the reference measure from before, the second term becomes a multiple
of the cross entropy, namely $- NM H(q,r)$.
Using Stirling's approximation, the first term becomes 
\begin{align}
    \sum_{i,j} - NMQ_{i,j} \log NMQ_{i,j} + NMQ_{i,j} + O(\log NMQ_{i,j})
\end{align}
On practical scales, the first two terms are the most significant.
Normalizing by the total $NM$ particles,
we note that this approaches the \emph{entropy} (in the statistics sense) of 
$q$ plus a constant
\begin{align}
    \sum_{i,j} -NMQ_{i,j} \log NMQ_{i,j} + NMQ_{i,j} &=
        NM H(p) - NM \log (NM) + NM
\end{align}
Thus our large $M$ approximation gives the posterior as 
\begin{align}
    \log p(Q) &= NM (H(q) - H(q,r)) + NM - NM \log (NM) \\
    &= NM KL(q | r) +  NM - NM \log (NM)
\end{align}
Once again, we see a power posterior. Note that there is another equivalent way we could
have reached this conclusion. First consider $NM$ particles
\emph{unconstrained} by the marginal condition, distributed independently such
that for each particle $P(X_0 = a_i, X_1 = b_j) \propto -\exp\left(-\frac 1 2 (a_i - b_j)^2 \right)$.
The distribution of these particles would correspond exactly to our
\emph{reference} measure from before. If we then \emph{condition} on the marginal
requirements we recover exactly the same posterior! In the nomenclature of
statistical mechanics, the reference measure corresponds to the \emph{energy}.

Either way we view it, in the limit of infinite $M$, the
distribution on $Q$ collapses to the delta distribution at the minimizer of the
$KL$ to the reference distribution subject to our marginal requirements. In
statistical mechanics we might interpret negative $KL$ as entropy. In other
words, we have discovered the principle of maximum entropy. We have also
discovered the concept of a Schrödinger Bridge. Another interesting point,
we make is to recall that the cross entropy corresponds to the cost function
from the Monge-Kantorovich problem. Thus this optimization problem corresponds
to an \emph{entropy-regularized} version of the Monge-Kantorovich problem,
where the variance of the Brownian motion changes the strength of
regularization.

\section{The Schrödinger Bridge Problem}
Armed with a deeper understanding of Brownian bridges, we are now ready to
state the Schrödinger bridge problem as the continous version of our previous
mixture of Brownian bridges problem. Briefly, given a Riemannian manifold
$\mathcal X$, the (static) Schrödinger bridge problem is to find the joint
distribution $Q$ on $\mathcal X \times \mathcal X$ which
\begin{enumerate}
    \item Satisfies the marginal constraints $X_0 \sim \mu$ and $X_1 \sim \nu$
    \item Minimizes the
KL-divergence to the reference Brownian motion $R$ on $\mathcal X \times \mathcal X$
\end{enumerate}
In the general case, the reference Brownian motion is given by
$R(x_0, x_1) \propto \exp\left(-\frac 1 2 (x_1 - x_0)^2 \right)$. The reference
is always a measure, but it may not be a probability measure. Once
a solution is found, the distribution on paths is just a mixture of
Brownian bridges with weights given by the transport plan.

\section{Schrödinger Bridge to Diffusion Models}
Before discussing how to actually compute the Schrödinger bridge, we touch on
why it might be useful for diffusion models. Traditional diffusion models
essentially consist of the following conceptual sequence
\begin{enumerate}
    \item Assume a form for the forward diffusion process with stationary
    distribution at the prior
    \item Run the forward process until it is close to stationary (i.e. the prior)
    \item Use samples from the forward process to \emph{learn} the backwards process.
    This amounts to learning the score
    \cite{ho2020denoisingdiffusionprobabilisticmodels}
    \cite{song2020generativemodelingestimatinggradients}
\end{enumerate}
We might be slightly uncomfortable with the second step above, since it relies
on us simply trusting that the diffusion process has been run long enough.
Solving the Schrödinger bridge problem helps us address this discomfort.
Instead of fixing the forward process and letting it run long enough to reach
stationarity, we can directly find the diffusion process that is closest to
our \emph{reference Brownian motion} while also satisfying the marginal
constraints. This means that we can now force the forward dynamics
to reach the prior \emph{in finite time}. This is the
first key idea in \cite{debortoli2023diffusionschrodingerbridgeapplications}.

\section{Solving the Schrödinger Bridge Problem}
We now return to the problem of actually solving the Schrödinger bridge problem.
We will focus again on a finite set as discussed in section
\ref{sec:many_particles}. First we will introduce the classic
Sinkhorn algorithm, before moving on to the implementation provided in 
\cite{debortoli2023diffusionschrodingerbridgeapplications}. In our context
the Sinkhorn algorithm is very simple. Letting $Q$ be transport matrix
normalized by the total number of particles. The Sinkhorn algorithm proceeds
as follows
\begin{enumerate}
    \item Initialize $Q$ at the reference distribution
    \item While not converged
    \begin{enumerate}
        \item Rescale the \textbf{rows} to satisfy the row marginal condition
        \item Rescale the \textbf{columns} to satisfy the column marginal conditon 
        \item Repeat
    \end{enumerate}
\end{enumerate}
It is a remarkable fact shown in \cite{Sinkhorn1964} that this algorithm
converges to the minimum entropy solution which satisfies the marginal
constraints! In fact it can be seen as a versin of coordinate ascent in the
\emph{dual} problem. We will not provide a proof of convergence here.

It may be helpful before moving forward to think of this algorithm in terms of
conditional distributions. Let rows correspond to
$X_0$ with target marginal $\mu$ and columns to $X_1$ with target marginal
$\nu$. The very first time we rescale the rows corresponds to computing the joint
distribution specified by $\mu$ and the \emph{reference dynamics}. This joint
distribution implies a \emph{backwards} conditional distribution $X_0|X_1$.
Crucially, if we now reweight the columns, the \emph{backwards} conditional distribution
is maintained and the $X_1$ marginal is adjusted. However, the forward
conditional $X_1 | X_0$ and $X_0$ marginal will now be perturbed. We can
repeat the process, reweighting the $X_0$ marginal and following the new
forward dynamics. The algorithm
will repeat until both marginal constraints are satisfied.

The Sinkhorn algorithm would be perfect if our distributions had finite support.
However, in diffusion modelling we are dealing with \emph{continuous priors},
\emph{high dimensions} and \emph{large datasets} which make a Sinkhorn
discretization infeasible. We could model the potentials using some flexible
parametric family, but finding an appropriate family is a non-trivial problem.
Another solution is to implicitly define the joint distribution by modelling
the \emph{diffusion} process itself. This is the key algorithmic innovation
introduced
in \cite{debortoli2023diffusionschrodingerbridgeapplications}.
At a high-level the algorithm is as follows
\begin{enumerate}
    \item Initialize at the data distribution and use the reference distribution
    for our forward dynamics
    \item Learn the backward dynamics by sampling from the data and propogating
    under the forward dynamics
    \item Learn the forward dynamics by sampling from the prior and
    propogating backwards under our learned backward dynamics
    \item Repeat
\end{enumerate}
This is exactly a continuous analog of the Sinkhorn algorithm, but where
conditional distributions are given by \emph{diffusion dynamics}.

\section{Simulation by MCMC}
In section \ref{sec:many_particles} we introduced a probability distribution
over transport plans in the finite case. We now introduce a
Markovian interacting particle system which admits this distribution as
a stationary distribution. The algorithm is described in Algorithm~\ref{alg:bitflip-mh}.
This algorithm is based on Metropolis-Hastings using a combination of
bit-masked swap proposals and random pairings. It alternates between a
\emph{y-swap phase} and an \emph{x-swap phase}: each phase sweeps through all
$N$ bit indices proposing swaps of the respective coordinate, then runs a
random-pairing shuffle of that same coordinate.  Because $\Delta E$ is symmetric
in $x$ and $y$, both swap types share the same MH acceptance ratio.  Stationarity is maintained
since each individual swap proposal is reversible (after MH correction), and
each phase preserves the empirical marginal of the coordinate being swapped. The
target is log concave \cite{léonard2013surveyschrodingerproblemconnections},
therefore multimodality is not an issue. An
interactive version of the simulation can be found on
\href{https://zach-lau.github.io/STAT547E-Project/code/particle_sim.html}{Github}
\footnote{\url{https://zach-lau.github.io/STAT547E-Project/code/particle_sim.html}}.

\begin{algorithm}[h]
\caption{Bit-Flip MH Kernel for Regularised OT}
\label{alg:bitflip-mh}
\begin{algorithmic}[1]
\Require $M = 2^N$ particles with positions $x_1,\ldots,x_M \sim r$ and
         $y$-values $y_1,\ldots,y_M$ with empirical marginal $c$,
         regularisation $\varepsilon > 0$
\State $\mathit{bit} \gets 0$,\quad $\mathit{phase} \gets \text{y-swap}$
\Loop
    \State $\mathit{mask} \gets 2^{\mathit{bit}}$
    \For{$i = 0, 1, \ldots, M-1$ such that $(i \gg \mathit{bit}) \mathbin{\&} 1 = 0$}
        \State $j \gets i \oplus \mathit{mask}$
        \State $\Delta E \gets (x_i - y_j)^2 + (x_j - y_i)^2 - (x_i - y_i)^2 - (x_j - y_j)^2$
        \If{$\Delta E \leq 0$ \textbf{or} $U \sim \mathrm{Uniform}(0,1) < e^{-\Delta E / \varepsilon}$}
            \If{$\mathit{phase} = \text{y-swap}$} Swap $y_i \leftrightarrow y_j$
            \Else\ Swap $x_i \leftrightarrow x_j$
            \EndIf
        \EndIf
    \EndFor
    \State $\mathit{bit} \gets (\mathit{bit} + 1) \bmod N$
    \If{$\mathit{bit} = 0$} \Comment{End of full $N$-bit cycle}
        \State Draw a uniformly random permutation $\sigma$ of $\{1,\ldots,M\}$
        \For{$k = 1, 3, 5, \ldots, M-1$} \Comment{Consecutive pairs}
            \If{$\mathit{phase} = \text{y-swap}$}
                \State Propose $y$-swap between $\sigma(k)$ and $\sigma(k+1)$ via MH
            \Else
                \State Propose $x$-swap between $\sigma(k)$ and $\sigma(k+1)$ via MH
            \EndIf
        \EndFor
        \State Toggle $\mathit{phase}$: y-swap $\leftrightarrow$ x-swap
    \EndIf
\EndLoop
\end{algorithmic}
\end{algorithm}

\section{Conclusion}
In this project we have built up a deeper understanding of the Schrödinger
bridge problem based on a finite particle approach. We worked up from Brownian
bridges, to interacting particle systems to demonstrate how the KL-divergence
naturally arises as a cost-function when considering diffusion problems with
constrained marginals. Furthermore, we introduced a simple MCMC algorithm to
simulate from the posteriors defined by applying the marginal constraints to
the reference dynamics in finite particle systems.

\bibliographystyle{plain}
\bibliography{refs}


\end{document}